A unified pseudo-$C_\ell$ framework

### 2018-09-08 doc/Paper/main.tex

        In the standard incarnation of the optimal quadratic estimator~\citep{1998PhRvD..57.2117B}, the measurement is produced on inverse covariance weighted data, namely
        \begin{equation}
          {\sf F_{ij}} B_j = \frac{1}{2}\va^\dagger \tilde{\sf C}^{-1}\,{\sf P}_i\,\tilde{\sf C}^{-1} \va -
          \frac{1}{2}\rm{Tr}\left[\tilde{\sf C}^{-1}{\sf P}_i\tilde{\sf C}^{-1}\,{\sf N} \right],
        \end{equation}
        where ${\sf F}$ is the Fisher matrix, the covariance matrix is modeled as a linear sum over band-power parameters $B_i$, ie.
        \begin{equation}\label{eq:bpws_qmv}
          \tilde{\sf C} = \sum_i {\sf P}_i\,B_i + {\sf N},
        \end{equation}
        and the last term in Eq. \ref{eq:bpws_qmv} corresponds to a bias of the convolved estimates due to the presence of noise. Thus, this optimal estimator corresponds to i) inverse-variance weighting the data, ii) calculating the spectrum of the weighted data using the response matrices ${\sf P}_i$ and iii) deconvolving these estimates using a mode-coupling matrix which in this particular case happens to be the same as Fisher matrix. The main computational challenge when implementing this estimator usually lies in storing the covariance matrix and its inverse (or computing it on the fly), particularly in order to estimate the Fisher matrix. This motivates the pseudo-$C_\ell$ estimator, which replaces these computationally very intensive steps with numerically more efficient ones at the expense of optimality.

@@

    \subsection{Contaminant deprojection}\label{ssec:maths.deproj}
      In a typical data-analysis case, we deal with contaminants that pollute the measured signal. Usually, we know something about these contaminants. As an example, in case of CMB analysis, we might have have a map of a given foreground at a different frequency which can be marginalized out. But even when we do not, we often have a tracer map of a given effect, for example, the map of dust might tell us something about systematic errors related to imperfect correction to dust reddening, or a map of co-added PSF fluctuations could be used to remove the possible correlations between measured ellipticities. Even if the dust map is non-linearly and perhaps stochastically related to the amount of reddening, as long as that relation is local in space, we get a linear contamination on large scales, very much like complex astrophysics of galaxy formation nevertheless reduces to linear biasing on very large scales (see Appendix \ref{app:lincont}). Thus, we generically deal with a linear model of the kind:
      \begin{equation}\label{eq:aobs}
        {\bf a}^{\rm obs}(\nv)={\bf a}^v+\sum_{i=1}^{N_a}\alpha_i{\bf f}^i(\nv),
      \end{equation}
      where the observed map ${\bf a}^{\rm obs}$ is composed of a masked signal map that receives contributions from $N_a$ known contaminants ${\bf f}^i$ and where the contamination coefficients $\alpha_i$ are unknown (note that we have implicitly applied the same mask to the contaminant templates ${\bf f}^i$). The deprojection of these contaminants from the data can be done and propagated into the estimated power spectrum in a natural way in the case of quadratic estimators \citep{1992ApJ...398..169R,2004PhRvD..69l3003S,2016MNRAS.456.2095E}, and the method can be extended to the pseudo-$C_\ell$ algorithm \citep{2017MNRAS.465.1847E}. The process is illustrated in Figure \ref{fig:lss_cont}. A best-fit value for the coefficients $\alpha_i$ can be found assuming \emph{uncorrelated weights} across the map\footnote{The assumption of uncorrelated weights is the key difference between pseudo-$C_\ell$ and optimal quadratic estimators \citep{2013MNRAS.435.1857L}. Note that, since the templates are already multiplied by the mask, the weights are not equal across the map, but correspond to the mask values in general.}
      \begin{align}\label{eq:deproj_coeff}
        &\hat{\alpha}_i=F_{ij}\int d\nv\,{\bf f}^{j\dag}(\nv)\,{\bf a}^{\rm obs}(\nv),\\\nonumber
        &({\sf F}^{-1})_{ij}\equiv\int d\nv\, {\bf f}^{i\dag}(\nv)\,{\bf f}^j(\nv),
      \end{align}
      where there is an implicit summation over the repeated index $j$ in Eq. \ref{eq:deproj_coeff} (which we will omit from now on). The clean map is then given by
      \begin{align}
        {\bf a}^c(\nv)
        &\equiv{\bf a}^{\rm obs}(\nv)-{\bf f}^i(\nv)F_{ij}\int d\nv'\,{\bf f}^{j\dag}(\nv')\,{\bf a}^{\rm obs}(\nv')
      \end{align}
      which, when substituting Eq. \ref{eq:aobs}, yields
      \begin{align}\label{eq:clean}
        {\bf a}^c(\nv)&={\bf a}^v(\nv)-{\bf f}^i(\nv)F_{ij}\int d\nv'\,{\bf f}^{j\dag}(\nv')\,{\bf a}^v(\nv').
      \end{align}      
      The harmonic coefficients of ${\bf a}^c$ are therefore given by
      \begin{equation}\label{eq:clean_f}
        {\bf a}^c_{\bf l}={\bf a}^v_{\bf l}-{\bf f}^i_{\bf l}\,F_{ij}\sum_{{\bf l}'}{\bf f}^{j\dag}_{{\bf l}'}{\bf a}^v_{{\bf l}'}
      \end{equation}

@@

      Secondly, estimating the deprojection coefficients (Eq. \ref{eq:deproj_coeff}) involves inverting the correlation matrix of all the contaminant templates. This will not be in general possible (e.g. if a set of templates are linearly related). The problem can however be solved self-consistently by using the Moore-Penrose pseudo-inverse \citep{2017MNRAS.465.1847E}. This is trivial for a symmetric and positive-definite matrix such as ${\sf F}$, and corresponds to setting the inverse of all of its zero eigenvalues (or all those below a given threshold) to zero.

>>>>

        In the standard incarnation of the optimal quadratic estimator~\citep{1998PhRvD..57.2117B}, the measurement is produced on inverse covariance weighted data, namely
        \begin{equation}
          {\sf F_{ij}} B_j = \frac{1}{2}\va^\dagger \tilde{\sf C}^{-1}\,{\sf P}_i\,\tilde{\sf C}^{-1} \va -
          \frac{1}{2}\rm{Tr}\left[\tilde{\sf C}^{-1}{\sf P}_i\tilde{\sf C}^{-1}\,{\sf N} \right],
        \end{equation}
        where ${\sf F}$ is the Fisher matrix, the covariance matrix is modeled as a linear sum over band-power parameters $B_i$, i.e.
        \begin{equation}\label{eq:bpws_qmv}
          \tilde{\sf C} = \sum_i {\sf P}_i\,B_i + {\sf N},
        \end{equation}
        (which defines the response matrices ${\sf P}_i$) and the last term in Eq. \ref{eq:bpws_qmv} corresponds to a bias of the convolved estimates due to the presence of noise. Thus, this optimal estimator corresponds to i) inverse-variance weighting the data, ii) calculating the spectrum of the weighted data using the response matrices and iii) deconvolving these estimates using a mode-coupling matrix which in this particular case happens to be the same as Fisher matrix. The main computational challenge when implementing this estimator usually lies in storing the covariance matrix and its inverse (or computing it on the fly), particularly in order to estimate the Fisher matrix. This motivates the pseudo-$C_\ell$ estimator, which replaces these computationally very intensive steps with numerically more efficient ones at the expense of optimality.

@@

    \subsection{Contaminant deprojection}\label{ssec:maths.deproj}
      In a typical data-analysis case, we deal with contaminants that pollute the measured signal. Usually, we know something about these contaminants. As an example, in case of CMB analysis, we might have a map of a given foreground at a different frequency which can be marginalized out. But even when we do not, we often have a tracer map of a given effect, for example, the map of dust might tell us something about systematic errors related to imperfect correction to dust reddening, or a map of co-added PSF fluctuations could be used to remove the possible correlations between measured ellipticities. Even if the dust map is non-linearly and perhaps stochastically related to the amount of reddening, as long as that relation is local in space, we get a linear contamination on large scales, very much like complex astrophysics of galaxy formation nevertheless reduces to linear biasing on very large scales (see Appendix \ref{app:lincont}). Thus, we generically deal with a linear model of the kind:
      \begin{equation}\label{eq:aobs}
        {\bf a}^{\rm obs}(\nv)={\bf a}^v+\sum_{i=1}^{N_a}\alpha_i{\bf f}^i(\nv),
      \end{equation}
      where the observed map ${\bf a}^{\rm obs}$ is composed of a masked signal map that receives contributions from $N_a$ known contaminants ${\bf f}^i$ and where the contamination coefficients $\alpha_i$ are unknown (note that we have implicitly applied the same mask to the contaminant templates ${\bf f}^i$). The deprojection of these contaminants from the data can be done and propagated into the estimated power spectrum in a natural way in the case of quadratic estimators \citep{1992ApJ...398..169R,2004PhRvD..69l3003S,2016MNRAS.456.2095E}, and the method can be extended to the pseudo-$C_\ell$ algorithm \citep{2017MNRAS.465.1847E}. The process is illustrated in Figure \ref{fig:lss_cont}. A best-fit value for the coefficients $\alpha_i$ can be found assuming \emph{uncorrelated weights} across the map\footnote{The assumption of uncorrelated weights is the key difference between pseudo-$C_\ell$ and optimal quadratic estimators \citep{2013MNRAS.435.1857L}. Note that, since the templates are already multiplied by the mask, the weights are not equal across the map, but correspond to the mask values in general.}
      \begin{align}\label{eq:deproj_coeff}
        &\hat{\alpha}_i=F_{ij}\int d\nv\,{\bf f}^{j\dag}(\nv)\,{\bf a}^{\rm obs}(\nv),\\\nonumber
        &({\sf F}^{-1})_{ij}\equiv\int d\nv\, {\bf f}^{i\dag}(\nv)\,{\bf f}^j(\nv),
      \end{align}
      where there is an implicit summation over the repeated index $j$ in Eq. \ref{eq:deproj_coeff} (which we will omit from now on). The clean map is then given by
      \begin{align}
        {\bf a}^c(\nv)
        &\equiv{\bf a}^{\rm obs}(\nv)-{\bf f}^i(\nv)F_{ij}\int d\nv'\,{\bf f}^{j\dag}(\nv')\,{\bf a}^{\rm obs}(\nv')
      \end{align}
      which, when substituting Eq. \ref{eq:aobs}, yields
      \begin{align}\label{eq:clean}
        {\bf a}^c(\nv)&={\bf a}^v(\nv)-{\bf f}^i(\nv)F_{ij}\int d\nv'\,{\bf f}^{j\dag}(\nv')\,{\bf a}^v(\nv').
      \end{align}      
      The harmonic coefficients of ${\bf a}^c$ are therefore given by
      \begin{equation}\label{eq:clean_f}
        {\bf a}^c_{\bf l}={\bf a}^v_{\bf l}-{\bf f}^i_{\bf l}\,F_{ij}\sum_{{\bf l}'}{\bf f}^{j\dag}_{{\bf l}'}{\bf a}^v_{{\bf l}'}
      \end{equation}

@@

      Secondly, estimating the deprojection coefficients (Eq. \ref{eq:deproj_coeff}) involves inverting the correlation matrix of all the contaminant templates. This will not be in general possible (e.g. if a set of templates are linearly related). The problem can however be solved self-consistently by using the Moore-Penrose pseudo-inverse \citep{2017MNRAS.465.1847E}. This is trivial for a symmetric and positive-definite matrix such as ${\sf F}$, and corresponds to setting the inverse of all of its zero eigenvalues (or all those below a given threshold) to zero\footnote{The choice of eigenvalue threshold in \nmt~is accessible to the user.}.
